---
title: "Predicting NBA shots"
output: rmarkdown::github_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

# Introduction
In this notebook, regression and machine learning techniques are used to predict
whether a basketball shot is succesful (hits the basket and scores a point) or not.
SPecifically, we'll look at all the basketball shots made during the 2014-2015 NBA 
season scraped from the NBA's API and provided on kaggle [(https://www.kaggle.com/dansbecker/nba-shot-logs/home)](https://www.kaggle.com/dansbecker/nba-shot-logs/home).

# Set up
```{r echo=T, results='hide'}
# Code which explores the NBA shot log data
# clear workspace
remove(list = ls())
# clear console
cat("\014")

# load packages
library(caret)
library(tidyverse)
library(skimr)
library(ggplot2)
library(tictoc)
library(corrplot)
library(pROC)
library(viridis) # for density plot
library(rpart.plot)

path_project <- "C:/Users/Mats Ole/Desktop/predicting_nba_shots/"
path_rel_data <- "data_input/"
name_data <- "shot_logs.csv"
```

# Data cleaning
First, let's clean and rename the data.
```{r}
df_raw <- read_csv(paste0(path_project, path_rel_data, name_data))

df_clean <- df_raw %>%
  transmute(game_id = as.factor(GAME_ID),
            matchup = as.factor(MATCHUP),
            home_game = case_when(LOCATION == "H" ~ TRUE,
                                  LOCATION == "A" ~ FALSE),
            win = case_when(W == "W" ~ TRUE,
                           W == "L" ~ FALSE),
            final_margin = abs(FINAL_MARGIN),
            shot_number = SHOT_NUMBER,
            period = PERIOD,
            game_clock = as.numeric(GAME_CLOCK),
            shot_clock = SHOT_CLOCK,
            dribbles = DRIBBLES,
            touch_time = TOUCH_TIME,
            shot_dist = SHOT_DIST,
            pts_type = as.factor(PTS_TYPE),
            shot_result = as.factor(SHOT_RESULT),
            closest_defender = as.factor(CLOSEST_DEFENDER),
            closest_defender_id = CLOSEST_DEFENDER_PLAYER_ID,
            closest_defender_dist = CLOSE_DEF_DIST,
            pts = PTS,
            player_name = as.factor(player_name),
            player_id = player_id) %>%
  drop_na() %>% # Here we're dropping about 6k or 0.4% of the observations.
  select(shot_result, everything())
```



# Data exploration
## Summary stats
```{r echo=T, results='asis'}
kable(skim(df_clean))
```
## Overall shot percentage
45.6% of the shots are succesful.
```{r echo=T}
mean(df_clean$shot_result == "made")
```

## Correlations among predictors
There are some expected correlations among the predictors: 

* As shot distance increases,
so does the space a defender gives the shooting player. 
* Number of dribbles taken by the player who makes the shots and the amount of
time the player has had the ball in his hands are naturally highly correlated.
* The number of the shot and the number of the game period/quarter are also expected to be correlated.
```{r echo=T}
correlations <- cor(df_clean %>% select_if(is.numeric))
corrplot(correlations, order = "hclust")
```

## Close and uncontested shots are more succesful
```{r echo=T}

ggplot(df_clean %>% 
         filter(shot_result == "made"), aes(x = shot_dist, 
                                            y = closest_defender_dist)) +
  stat_density2d(aes(fill = ..density..), contour = F, geom = 'tile') +
  scale_fill_viridis() + ggtitle("Density distribution of succesful shots") +
  coord_cartesian(xlim = c(0, 30), ylim = c(0, 20)) +
  geom_vline(xintercept = 22, linetype = "dotted", colour = "red") +
  geom_vline(xintercept = 23.75, linetype = "dotted", colour = "red") +
  geom_vline(xintercept = 23.75, linetype = "dotted", colour = "red") +
  annotate("text", x = 21, y = 15, 
           label = paste("Corner 3 points"), 
           size = 4, angle = 90, colour = "red") +
  annotate("text", x = 24.75, y = 15, 
           label=paste("Normal 3 points"), 
           size = 4, angle = 90, colour = "red")


ggplot(df_clean %>% filter(shot_result == "missed"), aes(x = shot_dist, y = closest_defender_dist)) +
  stat_density2d(aes(fill = ..density..), contour = F, geom = 'tile') +
  scale_fill_viridis() + ggtitle("Density distribution of unsuccesful shots") +
  coord_cartesian(xlim = c(0, 30), ylim = c(0, 20)) +
  geom_vline(xintercept = 22, linetype = "dotted", colour = "red") +
  geom_vline(xintercept = 23.75, linetype = "dotted", colour = "red") +
  geom_vline(xintercept = 23.75, linetype = "dotted", colour = "red") +
  annotate("text", x = 21, y = 15, 
           label = paste("Corner 3 points"), 
           size = 4, angle = 90, colour = "red") +
  annotate("text", x = 24.75, y = 15, 
           label=paste("Normal 3 points"), 
           size = 4, angle = 90, colour = "red")

```
